---
title: "Construction of foundation model ... Test and learn :)"
author: 
  - name: "Pierre Bady"
  - orcid: 0000-0003-1398-4689
  - email: "pierre.bad@unil.ch"
format: 
  html:
    theme: flatly
    toc: true
    toc-depth: 6
    toc-location: left-body
    number-sections: true
    number-depth: 6
    
editor: visual

execute: 
  warning: FALSE
  
Jupyter: /export/scratch3/gitProject/biomarkfinder/myenv
---

- **Contributor:** P.BADY
- **Modification Date:** `r date()`
- **Abstract:**

```         
License: GPL version 2 or newer
Copyright (C) 2009-2026  Pierre Bady

This program is free software/document; you can redistribute it and/or
modify it under the terms of the GNU General Public License
as published by the Free Software Foundation; either version 2
of the License, or (at your option) any later version.

This program/document is distributed in the hope that it will be useful,
but WITHOUT ANY WARRANTY; without even the implied warranty of
MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
GNU General Public License for more details.
```

```{r setup, include=FALSE}
# knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
library(rmarkdown)
library(utils)
options(encoding="UTF-8")
options(show.signif.stars=TRUE)
# non permanent command
load("/export/scratch3/gitProject/biomarkfinder/trunk/dev/.RData")
source("/export/scratch2/Rplus/Rutility.R")
library(xtable)
options(xtable.floating = FALSE)
options(xtable.timestamp = "")
library(xlsx)

# additional functions
.cleanPackage <- function(keeplib){
  auxi <- search()
  auxi <- auxi[!is.element(auxi,keeplib)]
  sapply(auxi,function(x) try(detach(x,character.only = TRUE,unload=TRUE)))
  R.utils::gcDLLs()
}
cleanPackage <- function(...){
  suppressPackageStartupMessages(.cleanPackage(...))
}
RequirePB <- function(...){
  suppressPackageStartupMessages(require(...))
}
SourcePB <- function(...){
  suppressPackageStartupMessages(source(...)) 
}
gg_color_hue <- function(n) {
  hues = seq(15, 375, length = n + 1)
  hcl(h = hues, l = 65, c = 100)[1:n]
}
  
### color gradient
cols <- colorRampPalette(rev(brewer.pal(n = 7, name ="RdYlBu")))(25)
```

# Motivations

# Virtual environment preparation

new alternative (2026-07-27): see https://quarto.org/docs/projects/virtual-environments.html

add the following command in a terminal (for station P3 ultra)

```         
cd /export/scratch3/gitProject/biomarkfinder
/export/scratch2/applications/anaconda3/bin/python -m venv myenv
source myenv/bin/activate
pip install pooch scanpy pandas numpy seaborn igraph leidenalg scipy scikit-learn datasets "transformers<=4.44.0" geneformer torch
cd /export/scratch3/gitProject/Geneformer
pip install .
```

installation of geneformer

```         
git lfs install
git clone https://huggingface.co/ctheodoris/Geneformer
cd Geneformer
pip install .
```

# Tokenization for RNA-seq with R

## data preparation
A classical idea is to use rank-based encoding where one sample (or cell) is a sentence and the tokens are genes (e.g. Geneformer,scGPT).



 * **Methodological note on the in silico perturbation strategy:** the approach of removing each gene from a cell's tokenized sequence and measuring the resulting shift in predicted probability (used in perturbation_scores_py311.py and run_biomarker_ranking.py) is inspired by the in silico deletion strategy described in the original Geneformer paper (Theodoris et al., 2023, "In silico perturbation" section) — this should be cited/credited as a derived method rather than the official implementation, since it is a simplified reimplementation for biomarker ranking purposes, not the model's source code.

## Two groups: Exemple for RNA-seq

Import count data from bulk RNAseq (https://zenodo.org/records/22207726)

```{r eval=FALSE}
annotated_combined <- read.delim("/export/scratch3/gitProject/biomarkfinder/trunk/data/annotated_combined.counts")
tabcount <- apply(annotated_combined[,!is.element(colnames(annotated_combined),c("id","symbol"))],2,function(x) tapply(x,annotated_combined$symbol,median))
saveRDS(tabcount,file="/export/scratch3/gitProject/biomarkfinder/trunk/data/tabcount.rds")
```

```{r eval=FALSE}
annotated_combined <- read.delim("/export/scratch3/gitProject/biomarkfinder/trunk/data/annotated_combined.counts")
rownames(annotated_combined) <- annotated_combined$id
tabcountEns <- annotated_combined[,!is.element(colnames(annotated_combined),c("id","symbol"))]
saveRDS(tabcountEns,file="/export/scratch3/gitProject/biomarkfinder/trunk/data/tabcountEns.rds")
```

```{r}
tabcount <- readRDS("/export/scratch3/gitProject/biomarkfinder/trunk/data/tabcountEns.rds")
dim(tabcount)
colnames(tabcount) <- gsub("[.]","_",colnames(tabcount))
colnames(tabcount) <- ifelse(colnames(tabcount)=="P670_036B","P670_036Bis",colnames(tabcount))
```

```{r}
library(readxl)
desc0 <- as.data.frame(read_excel("/export/scratch3/gitProject/biomarkfinder/trunk/data/TableS1_GRvolume.xlsx"))
rownames(desc0) <- desc0$ID
head(desc0)
dim(desc0)
```

filtering

```{r}
library(edgeR)
intersectx <- intersect(colnames(tabcount),rownames(desc0))
desc0 <- desc0[intersectx,]
tabcount <- tabcount[,rownames(desc0)]

###
DGEcount0 <-DGEList(counts=tabcount) #creat DGElist object

#data preparation and filtering for genes with very low counts
#the experiment has 3 replicates for each condition. We keep genes with at least 2 cpm in at least 3 samples
# keep <- rowSums(cpm(DGEcount0)>2 )>=3
cpm <- edgeR::cpm
## updated 07/06/2023 => more relax on the selection of the gene
keep <- rowSums(cpm(DGEcount0)>2 )>=1
table(keep)
DGEcount0 <-DGEcount0[keep, , keep.lib.sizes = F]

### need to examine more en details human|mice|treatment|tuba-tumour
design1 <-model.matrix( ~ 1, data=desc0)
colnames(design1)

#Calculate normalization factors to scale the raw library sizes
#counthuman0 <- counthuman
DGEcount0 <-calcNormFactors(DGEcount0) 

#takes really long time. includes common and tagwise dispersion. tagwise is recommende for differential expression analysis 
DGEcount0 <- estimateDisp(DGEcount0, design=design1, robust = TRUE) 
saveRDS(DGEcount0,file="DGEcount0.rds")
#save(counthuman0,file="counthuman0.rda")
saveRDS(desc0,file="desc0.rds")
```


load R functions for the preparation of the tokenization
```{r}
source("/export/scratch3/gitProject/biomarkfinder/trunk/code/tokenize.R")
```

Each cell or sample is converted into an ordered vector of gene identifiers (ranked by descending relative expression) and then mapped to token identifiers (integers) using a vocabulary, before being fed into a model based on the Transformer architecture. This step is typically performed in Python, as the model training itself is rarely carried out in R. R is well-suited for constructing the vocabulary and ordered sequences; the data are then exported in JSON or Parquet format for use with `transformers` (PyTorch).

```{r}
# usage
DGCcount <- Matrix(DGEcount0$counts, sparse = T )
tokens_list <- tokenize_expression(DGCcount)
head(tokens_list[[1]])  # ranked gene-ID sequence for cell 1
```

construction Json object

```{r}
library(jsonlite)
labels <- desc0$subtype

# export as JSON lines: one record per cell, ready for a HF `datasets.Dataset`
records <- Map(function(tok, lab) list(input_tokens = tok, label = lab),
               tokens_list, labels)
write(sapply(records, toJSON, auto_unbox = TRUE), file = "tokens_labeled.jsonl")
```

problem with geneformer: need to downgrade transformers

```         
pip install "transformers<=4.44.0"
```

You can fix this error by downgrading your transformers package to version 4.44.0 or older, where SpecialTokensMixin was still exposed directly

preparation of the model in python

```{python, eval=FALSE}
from datasets import load_dataset
from transformers import AutoModelForSequenceClassification, Trainer, TrainingArguments
#from geneformer import TranscriptomeTokenizer  # or scGPT equivalent
import torch
import geneformer

# gene tokens -> integer IDs using the model's pretrained vocabulary
dataset = load_dataset("json", data_files="tokens_labeled.jsonl")["train"]
label2id = {"non-codel": 0, "codel": 1}
dataset = dataset.map(lambda x: {"label": label2id[x["label"]]})

model = AutoModelForSequenceClassification.from_pretrained(
    "ctheodoris/Geneformer", num_labels=2
)

args = TrainingArguments(
    output_dir="geneformer_biomarker_ft",
    per_device_train_batch_size=16,
    num_train_epochs=5,
    learning_rate=5e-5 #,evaluation_strategy="epoch"
)

trainer = Trainer(model=model, args=args, train_dataset=dataset)
trainer.train()
```

```{python}
# """
# Diagnostic isolé — à exécuter seul via reticulate::py_run_file(), séparément
# du script d'entraînement complet. Ne fait aucun entraînement, juste des
# vérifications sur tokens_labeled.jsonl et le vocabulaire Geneformer.
# """

import json
from transformers import AutoTokenizer

# 1. Inspecter le fichier brut, sans rien via `datasets`
with open("tokens_labeled.jsonl") as f:
    lines = [json.loads(l) for l in f]

print(f"Nombre total de lignes dans tokens_labeled.jsonl : {len(lines)}")
print(f"Clés du premier enregistrement : {list(lines[0].keys())}")
print(f"Type de 'input_tokens' du 1er enregistrement : {type(lines[0]['input_tokens'])}")
print(f"5 premiers tokens du 1er enregistrement : {lines[0]['input_tokens'][:5]}")
print(f"Nombre de tokens dans le 1er enregistrement : {len(lines[0]['input_tokens'])}")
print(f"Valeur de 'label' du 1er enregistrement : {lines[0]['label']!r}")

labels_seen = set(l["label"] for l in lines)
print(f"Valeurs uniques de 'label' dans tout le fichier : {labels_seen}")

# 2. Charger le vocabulaire Geneformer et vérifier le format exact attendu
tokenizer = AutoTokenizer.from_pretrained("ctheodoris/Geneformer")
vocab = set(tokenizer.get_vocab().keys())
print(f"Taille du vocabulaire Geneformer : {len(vocab)}")
print(f"5 exemples d'entrées du vocabulaire : {list(vocab)[:5]}")

# 3. Vérifier si un suffixe de version (.xx) casse le matching
sample = lines[0]["input_tokens"][:20]
exact_overlap = sum(1 for t in sample if t in vocab)
stripped_overlap = sum(1 for t in sample if t.split(".")[0] in vocab)
print(f"Overlap exact (avec suffixe éventuel) sur 20 tokens : {exact_overlap}/20")
print(f"Overlap après suppression du suffixe '.xx' sur 20 tokens : {stripped_overlap}/20")

# 4. Overlap sur l'ensemble du fichier (pas juste le 1er enregistrement)
all_tokens = set()
for l in lines:
    all_tokens.update(l["input_tokens"])
total_overlap = len(all_tokens & vocab)
print(f"Nombre de gènes uniques dans le fichier : {len(all_tokens)}")
print(f"Nombre de gènes uniques présents dans le vocabulaire Geneformer : {total_overlap}")
```

```{python}
# Diagnostic : localiser token_dictionary.pkl SANS exécuter geneformer/__init__.py
# (qui plante à cause d'une incompatibilité de version avec transformers).
# importlib.util.find_spec() localise le package sans l'importer/exécuter.

import importlib.util
import os
 
spec = importlib.util.find_spec("geneformer")
if spec is None or spec.origin is None:
    raise RuntimeError("Le package 'geneformer' est introuvable dans l'environnement Python actif.")
 
pkg_dir = os.path.dirname(spec.origin)
print(f"Dossier d'installation de geneformer (sans l'importer) : {pkg_dir}")
 
print("\nFichiers .pkl ou contenant 'token'/'dict' dans ce dossier (récursif) :")
found = []
for root, dirs, files in os.walk(pkg_dir):
    for f in files:
        if f.endswith(".pkl") or "token" in f.lower() or "dict" in f.lower():
            full_path = os.path.join(root, f)
            found.append(full_path)
            print(full_path)
 
if not found:
    print("Aucun fichier trouvé — le dictionnaire est peut-être téléchargé séparément "
          "(pas inclus dans le package pip) ou porte un autre nom.")
 

```

```{r}
reticulate::py_run_file("/export/scratch3/gitProject/biomarkfinder/trunk/code/finetune_geneformer_py311.py")
```

next step: integration des deux scripts python (le code n'aime pas être décomposé)

```{r}
reticulate::py_run_file("/export/scratch3/gitProject/biomarkfinder/trunk/code/run_biomarker_ranking.py")
```


simplifyVector = FALSE évite que jsonlite ne fusionne les paires [gene_id, effect] (chaîne + nombre) en une matrice de caractères, ce qui forcerait effect en texte. Le code ci-dessus reconstruit un data.frame propre avec effect en numérique, trié dans le même ordre que dans le fichier (donc déjà par effet décroissant).

```{r}
library(jsonlite)
biomarkers <- fromJSON("top_biomarkers.json", simplifyVector = FALSE)
biomarkers_df <- do.call(rbind, lapply(biomarkers, function(x) {
  data.frame(gene_id = x[[1]], effect = as.numeric(x[[2]]))
}))

head(biomarkers_df, 20)
```


in construction

```{r eval=FALSE}
library(edgeR)   # or DESeq2

top_genes <- head(biomarkers_df, 20)$gene_id  # from top_biomarkers above

dge <- DGEList(counts = DGEcount0$counts[top_genes, ], group = desc0$subtype)
dge <- calcNormFactors(dge)
design <- model.matrix(~ desc0$subtype)
dge <- estimateDisp(dge, design)
fit <- glmQLFit(dge, design)
qlf <- glmQLFTest(fit, coef = 2)
topTags(qlf, n = length(top_genes))
```

## Multi-class

## Regression

```{python}
import numpy as np
import pandas as pd
from sklearn.decomposition import PCA
from sklearn.linear_model import Ridge
```
