---
title: "biomarkfinder"
author: 
  - name: "Pierre Bady"
  - orcid: 0000-0003-1398-4689
  - email: "pierre.bad@unil.ch"
format: 
  html:
    theme: flatly
    toc: true
    toc-depth: 6
    toc-location: left-body
    number-sections: true
    number-depth: 6
    
editor: visual

execute: 
  warning: FALSE
  
Jupyter: /export/scratch3/gitProject/biomarkfinder/trunk/myenv
---

- **Contributor:** P.BADY
- **Modification Date:** `r date()`
- **Abstract:**

```         
License: GPL version 2 or newer
Copyright (C) 2009-2026  Pierre Bady

This program is free software/document; you can redistribute it and/or
modify it under the terms of the GNU General Public License
as published by the Free Software Foundation; either version 2
of the License, or (at your option) any later version.

This program/document is distributed in the hope that it will be useful,
but WITHOUT ANY WARRANTY; without even the implied warranty of
MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
GNU General Public License for more details.
```

```{r setup, include=FALSE}
# knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
library(rmarkdown)
library(utils)
options(encoding="UTF-8")
options(show.signif.stars=TRUE)
# non permanent command
load("/export/scratch3/gitProject/biomarkfinder/trunk/dev/.RData")
source("/export/scratch2/Rplus/Rutility.R")
library(xtable)
options(xtable.floating = FALSE)
options(xtable.timestamp = "")
library(xlsx)

# additional functions
.cleanPackage <- function(keeplib){
  auxi <- search()
  auxi <- auxi[!is.element(auxi,keeplib)]
  sapply(auxi,function(x) try(detach(x,character.only = TRUE,unload=TRUE)))
  R.utils::gcDLLs()
}
cleanPackage <- function(...){
  suppressPackageStartupMessages(.cleanPackage(...))
}
RequirePB <- function(...){
  suppressPackageStartupMessages(require(...))
}
SourcePB <- function(...){
  suppressPackageStartupMessages(source(...)) 
}
gg_color_hue <- function(n) {
  hues = seq(15, 375, length = n + 1)
  hcl(h = hues, l = 65, c = 100)[1:n]
}
### color gradient
cols <- colorRampPalette(rev(brewer.pal(n = 7, name ="RdYlBu")))(25)
```

# Motivations

# Virtual environment preparation

new alternative (2026-07-27): see https://quarto.org/docs/projects/virtual-environments.html

add the following command in a terminal (for station P3 ultra)

```         
cd /export/scratch3/gitProject/biomarkfinder
/export/scratch2/applications/anaconda3/bin/python -m venv myenv
source myenv/bin/activate
pip install pooch scanpy pandas numpy seaborn igraph leidenalg scipy scikit-learn 
```





# Tokenization for RNA-seq with R

## data preparation
A classical idea is to use rank-based encoding where one sample (or cell) is a sentence and the tokens are genes (e.g. Geneformer,scGPT).

### Exemple for 
Import count data from bulk RNAseq (https://zenodo.org/records/22207726)
```{r}
annotated_combined <- read.delim("/export/scratch3/gitProject/biomarkfinder/trunk/data/annotated_combined.counts")
tabcount <- apply(annotated_combined[,!is.element(colnames(annotated_combined),c("id","symbol"))],2,function(x) tapply(x,annotated_combined$symbol,median))
saveRDS(tabcount,file="/export/scratch3/gitProject/biomarkfinder/trunk/data/tabcount.rds")
```


```{r}
tabcount <- readRDS("/export/scratch3/gitProject/biomarkfinder/trunk/data/tabcount.rds")
dim(tabcount)
colnames(tabcount) <- gsub("[.]","_",colnames(tabcount))

```


```{r}
source("/export/scratch3/gitProject/biomarkfinder/trunk/code/tokenize.R")
```


Each cell or sample is converted into an ordered vector of gene identifiers (ranked by descending relative expression) and then mapped to token identifiers (integers) using a vocabulary, before being fed into a model based on the Transformer architecture. This step is typically performed in Python, as the model training itself is rarely carried out in R. R is well-suited for constructing the vocabulary and ordered sequences; the data are then exported in JSON or Parquet format for use with `transformers` (PyTorch).




## Two groups

## Multi-class

## Regression



```{python}
import numpy as np
import pandas as pd
from sklearn.decomposition import PCA
from sklearn.linear_model import Ridge
```

